---
title: "Module 3 Assignment 1"
output: pdf_document
date: "2022-10-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assignment Details

### Purpose

The goal of this assignment is to assess your ability to produce and interpret histograms and scatter plots using `ggplot2`.

### Task

Write R code (using `ggplot2`, specifically) which produces the correct answers and correctly interpret the plots produced.

### Criteria for Success

-   Code is within the provided code chunks
-   Code is commented with brief descriptions of what the code does
-   Code chunks run without errors
-   Code produces the correct result
    -   Code that produces the correct answer will receive full credit
    -   Code attempts with logical direction will receive partial credit
-   Written answers address the questions in sufficient detail

### Due Date

November 8 at midnight MST

# Assignment Questions

For this assignment, we are going to be making plots! Specifically, we are going to be reproducing plots that we made in base R with `ggplot2`. If you want a refresher about the data and plots we are working with, take a gander at Module2_Assignment 1.

As before, we are going to use the data set called `penguins` from the `palmerpenguins` package.

Most of the code you will need to complete this assignment is code we used in the first 3 lessons of this module.

### Data

1.  Load both the `palmerpenguins` package and the `tidyverse` package into the work space. (2 points)

```{r}
library(tidyverse)
library(palmerpenguins)
```

When we use data from a data package, it doesn't automatically show up in our environment. Run this code chunk so it does show up in the environment.

```{r}
penguins <- penguins
```

2.  Use the `head()` function to refresh yourself on the `penguins` data frame. (1 points)

```{r}
head(penguins)
```

### Histogram (or Density Plot)

3.  Use `ggplot2` to make a histogram of the body mass column for all species combined (AKA don't set color or fill yet). You can create either a true histogram or a density plot, whichever you prefer. (2 points)

    *Note: it will produce a warning saying it removed some rows. That's fine!*

```{r}
ggplot(penguins, aes(body_mass_g)) +
  geom_density()

# OR 
ggplot(penguins, aes(body_mass_g)) +
  geom_histogram()

```

4.  Now, let's produce the same plot (histogram or density plot of the body mass column) but instead of a plot of all species combined, set the `color` and `fill` arguments to be determined by the species of penguin. Change the transparency of the fill (`alpha`) so we can see all three species. (2 points)

    (Hint: if you make a histogram, you'll want to set `position = "identity"` so you can see the histograms for all 3 species)

```{r}
ggplot(penguins, aes(body_mass_g, color = species, fill = species)) +
  geom_density(alpha = 0.5)

# OR #

ggplot(penguins, aes(body_mass_g, color = species, fill = species)) +
  geom_histogram(alpha = 0.5, position = "identity")

```

5.  In 2-3 sentences, describe what the histogram is telling you. How are the two histograms from questions 3 and 4 different? What different bits of information can you glean from presenting the histogram these two different ways? I'm not necessarily looking for technical answers, but I want you to practice interpreting what histograms are telling you. (3 points)

*Answer:*
Generally, did they interpret the histogram more-or-less correctly? Stuff like which mean is likely higher or lower, spread of the data, overlap of the data, etc.

6.  Let's spruce up the plot from question 4 a bit. (3 points)

Make the following changes:

-   edit the x-axis and y-axis labels to be capitalized and easier to read
-   capitalize the legend title ("Species" instead of "species")
-   choose a pre-programmed theme for your plot; I recommend `theme_bw()` or `theme_classic()`, but you can choose whichever one you like, as long as the axes titles and legend remain!

```{r}
ggplot(penguins, aes(body_mass_g, color = species, fill = species)) +
  geom_density(alpha = 0.5) +
  labs(x = "Body Mass (g)",
       y = "Count",
       color = "Species",
       fill = "Species") +
  theme_classic() # any theme is fine

# also fine if they use the xlab, ylab, scale_color_discrete, and scale_fill_discrete functions to relabel stuff
```

### Scatter Plot

7.  Make a scatter plot with flipper length on the x-axis (horizontal) and bill length on the y-axis (vertical) for all penguins, regardless of species. (2 points)

```{r}
ggplot(penguins, aes(flipper_length_mm, bill_length_mm)) +
  geom_point()
```

8.  Now, let's make this plot shine. Make the following edits. (3 point)

-   show a different marker color depending on the penguin species.
-   edit the x-axis and y-axis labels to be capitalized and easier to read
-   capitalize the legend title ("Species" instead of "species")
-   choose a pre-programmed theme for your plot; I recommend `theme_bw()` or `theme_classic()`, but you can choose whichever one you like, as long as the axes titles and legend remain!

```{r}
ggplot(penguins, aes(flipper_length_mm, bill_length_mm, color = species)) +
  geom_point() + 
  labs(x = "Flipper Length (mm)",
       y = "Bill Length (mm)",
       color = "Species") +
  theme_classic() # any theme is fine

# also fine if they use the xlab, ylab, scale_color_discrete, and scale_fill_discrete functions to relabel stuff
```

9.  Write 1-2 sentences discussing why including color based on species is important, based on the two plots above. (2 points)

*Answer:* something along the lines of it being clear that different species fall in clear groups
